Unifying Gompertzian growth with the communicable disease spreading paradigm

### 2022-04-29 main.tex
remove redundant folder

However, instead of showing logistic-like growth, observed cumulative mortality in the initial period March-April 2020 exhibits almost perfect resemblance to Gompertzian growth \citep{Gompertz1825} where the log-transformed cumulative mortality, or log-mortality for short, is exponentially \emph{decreasing} in time,
\begin{equation}
\label{eq:GompertzODE}
\frac{d}{dt}\ln{Y(t)} = -\beta\ln{\frac{Y(t)}{\tilde{Y}}} + \nu,
\end{equation} 
% Redundant definitions for full script
with constants $\tilde{Y}$, $\beta$, and $\nu$, and whose solution is given as
\begin{equation}
\label{eq:gomp_solution}
Y(t) = Y_\infty \left(\frac{Y_0}{Y_\infty}\right)^{e^{-\beta t}}
\end{equation}
$Y_{0}=Y(t = 0)$, $Y_{\infty}=Y(t\rightarrow \infty)=\tilde{Y}e^{\nu/\beta}$.

@@

With the full 3-parameter single ODE of the diseased group in the SIRD model, one can obtain a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity. 
Meanwhile, the Gompertz growth model can explain the initial observations with only two parameters obtainable through linear regression, while the third parameter accounts for the discrepancy between the final observation, $Y_{max}$, and $Y_{\infty}$, an effect which is manifested as a tapering of the initial straight line in the double-log domain. 
Thus, through linear regression estimates, the Gompertz model avoids the possibility of non-identifiability issues and unstable optima \citep{roda2020difficult}.
Comparisons between the two models and the inevitable concavity of the SIRD model can be seen when fitting observations of diseased attributed to the Coronavirus in a variety of countries (Fig. \ref{fig:Gompertz_vs_SIRD_lnln_ymax} and Supplemental Information).

@@

\insertPdfFig{Gompertz_vs_SIRD_lnln_ymax}{Cumulative number of diseased, transformed by $g(Y)=\ln{\ln{({Y_{max} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD with a Gompertz model. Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text after normalizing, although the Gompertz curve can also be obtained with a simple linear fit using (\ref{eq:GOMP_D}). The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm. Each plot is annotated with an inferred transmission rate in the SIRD model, assuming bi-normally distributed parameter estimates. Fitting is done with Python-Scipy and a 6-day windowed average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}

@@

\begin{subequations}
\begin{align}
\dot{F} = \frac{d}{dt}F(t) & = \beta F(t) (1-F(t)) \quad \text{Logistic}\label{eq:compareLog}\\ 
\frac{d}{dt}\ln{(F(t))} & = -\beta \ln{(F(t)}) \quad\quad\text{Gompertz}\label{eq:compareGom},
\end{align}
\end{subequations}
where $F$ is here deterministic.

@@

In contrast, as shown in (\ref{eq:microscopic}), the Gompertz model is implied by a multiplicative stochastic perturbation and thus implies correlated mortality growth by application of the central limit theorem in the log-domain, viz.
\begin{equation}
F(t) = \exp{\left[\frac{1}{N}\sum_i^N z_i(t)\right]} = \left[\prod_i^N f_i(t)\right]^{1/N},
\end{equation}
where $z_i$ is a random variable as in \ref{eq:microscopic} while $f_i$ is here deterministic. 
Under this model, correlation between entities is present at all orders in the original domain\footnote{It is illuminating to at this point compare with Gompertz' Law of Mortality, $\frac{d}{dt}\ln{(1 - F(t))} = -b$, for $t$ more than ~25 years, which yields a naturally uncorrelated macroscopic curve $1 - F(t) = \exp{(-bt)}$ \citep{shklovskii2005simple}. 
In our context, the uncorrelated feature emerges since the force of mortality is not a function of the growth itself, as we see in the text, but rather of time.}.

@@

\begin{table}[H]
\centering
\caption{Estimated values for the SIRD model fitted to various countries' observations normalized by the last observation, $Y_{max}$, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_ymax} using notation from \ref{eq:SIRD_D}. Note that due to normalization values will be different from \citet{carletti2020covid}.}
\begin{tabular}{llll}
\toprule
{} &           $\eta\times 10$ &      $\xi$ &            $\kappa\times 10^{2}$ \\
\midrule
Belgium        &    1.0(1) &  2.9(3) &   9.3(1) \\
Portugal       &   0.55(4) &  4.2(4) &   4.5(4) \\
Austria        &    1.3(3) &  2.3(4) &    11(2) \\
Switzerland    &   1.24(1) &  2.4(2) &  11.0(7) \\
Sweden         &    0.5(2) &    3(1) &     3(2) \\
Italy          &   0.43(1) &  5.6(2) &   3.7(1) \\
France         &   0.88(7) &  3.4(3) &   8.1(6) \\
Spain          &  0.525(1) &  6.2(2) &   4.8(1) \\
Germany        &   0.78(8) &  3.0(3) &   6.5(7) \\
Denmark        &   0.50(4) &  5.2(6) &   4.2(4) \\
United Kingdom &   0.54(2) &  4.7(2) &   4.3(2) \\
Netherlands    &   0.51(3) &  5.0(4) &   4.3(3) \\
Turkey         &   0.84(9) &  2.8(3) &   6.9(8) \\
Iran           &  0.322(7) &  7.2(3) &  2.51(9) \\
US             &  0.339(8) &  7.0(3) &   1.8(1) \\
China          &    3.8(9) &  1.2(2) &    26(4) \\
\bottomrule
\end{tabular}

@@

\begin{table}[H]
\centering
\caption{Estimated values for the Gompertz model fitted to various countries' observations normalized by the last observation, $Y_{max}$, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_ymax} using notation from \ref{eq:GompertzODE}.}
\begin{tabular}{llll}
\toprule
{} &             $\nu\times 10$ &           $\beta\times 10^{2}$ &        $\tilde{Y}\times 10^{2}$ \\
\midrule
Belgium        &   2.74(1) &   8.1(2) &  3.6(3) \\
Portugal       &  2.242(6) &  6.23(9) &  3.1(1) \\
Austria        &   2.74(1) &   8.1(2) &  3.6(2) \\
Switzerland    &  2.740(7) &   8.1(1) &  3.6(2) \\
Sweden         &   1.73(3) &   4.3(3) &  2.6(4) \\
Italy          &  2.107(6) &  5.88(9) &  3.0(1) \\
France         &  2.752(1) &   8.2(2) &  3.6(2) \\
Spain          &  2.556(8) &   7.5(1) &  3.4(2) \\
Germany        &  2.335(7) &   6.5(1) &  3.2(1) \\
Denmark        &   2.32(1) &   6.6(2) &  3.2(3) \\
United Kingdom &  2.333(6) &  6.54(9) &  3.2(1) \\
Netherlands    &  2.311(7) &   6.5(1) &  3.2(1) \\
Turkey         &  2.306(7) &   6.4(1) &  3.2(1) \\
Iran           &  1.862(7) &  5.04(9) &  2.7(1) \\
US             &  2.014(8) &   5.3(1) &  2.9(1) \\
China          &  3.155(8) &   9.4(2) &  3.6(2) \\
\bottomrule
\end{tabular}

>>>>

However, instead of showing logistic-like growth, observed cumulative mortality in the initial period March-April 2020 exhibits almost perfect resemblance to Gompertzian growth \citep{Gompertz1825} where the log-transformed cumulative mortality, or log-mortality for short, is exponentially \emph{decreasing} in time,
\begin{equation}
\label{eq:GompertzODE}
\frac{d}{dt}\ln{Y(t)} = -\beta\ln{\frac{Y(t)}{\tilde{Y}}} + \nu,
\end{equation} 
% Redundant definitions for full script
with constants $\tilde{Y}$, $\beta$, and $\nu$, and whose solution is given as
\begin{equation}
\label{eq:gomp_solution}
Y(t) = Y_\infty \left(\frac{Y_0}{Y_\infty}\right)^{e^{-\beta t}},
\end{equation}
where $Y_{0}=Y(t = 0)$ and $Y_{\infty}=Y(t\rightarrow \infty)=\tilde{Y}e^{\nu/\beta}$.

@@

With the full 3-parameter single ODE of the diseased group in the SIRD model, one can obtain a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity. 
Meanwhile, the Gompertz growth model can explain the initial observations with only two parameters obtainable through linear regression, while the third parameter accounts for the discrepancy between the final observation, $Y_{max}$, and $Y_{\infty}$ which is a parameter.
Thus, through linear regression estimates, the Gompertz model avoids the possibility of non-identifiability issues and unstable optima \citep{roda2020difficult}.
Comparisons between the two models and the inevitable concavity of the SIRD model can be seen when fitting observations of diseased attributed to the Coronavirus in a variety of countries (Fig. \ref{fig:Gompertz_vs_SIRD_lnln_infty} and Supplemental Information).

@@

\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of diseased, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD with a Gompertz model. Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text. Although the Gompertz curve can also be obtained with a simple linear fit using (\ref{eq:GOMP_D}), it is here obtained using non-linear least squares to put both models on equal footing. The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution. Each plot is annotated with an inferred transmission rate in the SIRD model, assuming bi-normally distributed parameter estimates. Fitting is done with Python-Scipy and a 6-day windowed average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}

@@

\begin{subequations}
\begin{align}
\dot{F} = \frac{d}{dt}F(t) & = \beta F(t) (1-F(t)) \quad \text{Logistic}\label{eq:compareLog}\\ 
\frac{d}{dt}\ln{(F(t))} & = -\beta \ln{(F(t)}) \quad\quad\text{Gompertz}\label{eq:compareGom},
\end{align}
\end{subequations}
where $F$ is in this section treated as deterministic.

@@

In contrast, as shown in (\ref{eq:microscopic}), the Gompertz model is implied by a multiplicative stochastic perturbation and thus implies correlated mortality growth by application of the central limit theorem in the log-domain and the continuous mapping theorem, viz.
\begin{equation}
\exp{\left[\frac{1}{N}\sum_i^N z_i(t)\right]} \xrightarrow{a.s.} \left[\prod_i^N f_i(t)\right]^{1/N} = F(t),
\end{equation}
where $z_i$ is a random variable as in \ref{eq:microscopic} while $f_i$ is here deterministic. 
Under this model, correlation between entities is present at all orders in the original domain\footnote{It is illuminating to at this point compare with Gompertz' Law of Mortality, $\frac{d}{dt}\ln{(1 - F(t))} = -b$, for $t$ more than ~25 years, which yields a naturally uncorrelated macroscopic curve $1 - F(t) = \exp{(-bt)}$ \citep{shklovskii2005simple}. 
In our context, the uncorrelated feature emerges since the force of mortality is not a function of the growth itself, as we see in the text, but rather of time.}.

@@

\begin{table}[H]
\centering
\caption{Estimated values for the SIRD model fitted to various countries' observations normalized by the last observation, $Y_{max}$, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:SIRD_D}. Note that due to normalization values will be different from \citet{carletti2020covid}.}
\begin{tabular}{llll}
\toprule
{} &           $\eta\times 10$ &      $\xi$ &            $\kappa\times 10^{2}$ \\
\midrule
Belgium        &    1.0(1) &  2.9(3) &   9.3(1) \\
Portugal       &   0.55(4) &  4.2(4) &   4.5(4) \\
Austria        &    1.3(3) &  2.3(4) &    11(2) \\
Switzerland    &   1.24(1) &  2.4(2) &  11.0(7) \\
Sweden         &    0.5(2) &    3(1) &     3(2) \\
Italy          &   0.43(1) &  5.6(2) &   3.7(1) \\
France         &   0.88(7) &  3.4(3) &   8.1(6) \\
Spain          &  0.525(1) &  6.2(2) &   4.8(1) \\
Germany        &   0.78(8) &  3.0(3) &   6.5(7) \\
Denmark        &   0.50(4) &  5.2(6) &   4.2(4) \\
United Kingdom &   0.54(2) &  4.7(2) &   4.3(2) \\
Netherlands    &   0.51(3) &  5.0(4) &   4.3(3) \\
Turkey         &   0.84(9) &  2.8(3) &   6.9(8) \\
Iran           &  0.322(7) &  7.2(3) &  2.51(9) \\
US             &  0.339(8) &  7.0(3) &   1.8(1) \\
China          &    3.8(9) &  1.2(2) &    26(4) \\
\bottomrule
\end{tabular}

@@

\begin{table}[H]
\centering
\caption{Estimated values for the Gompertz model fitted to various countries' observations normalized by the last observation, $Y_{max}$, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:GompertzODE}.}
\begin{tabular}{llll}
\toprule
{} &             $\nu\times 10$ &           $\beta\times 10^{2}$ &        $\tilde{Y}\times 10^{2}$ \\
\midrule
Belgium        &   2.74(1) &   8.1(2) &  3.6(3) \\
Portugal       &  2.242(6) &  6.23(9) &  3.1(1) \\
Austria        &   2.74(1) &   8.1(2) &  3.6(2) \\
Switzerland    &  2.740(7) &   8.1(1) &  3.6(2) \\
Sweden         &   1.73(3) &   4.3(3) &  2.6(4) \\
Italy          &  2.107(6) &  5.88(9) &  3.0(1) \\
France         &  2.752(1) &   8.2(2) &  3.6(2) \\
Spain          &  2.556(8) &   7.5(1) &  3.4(2) \\
Germany        &  2.335(7) &   6.5(1) &  3.2(1) \\
Denmark        &   2.32(1) &   6.6(2) &  3.2(3) \\
United Kingdom &  2.333(6) &  6.54(9) &  3.2(1) \\
Netherlands    &  2.311(7) &   6.5(1) &  3.2(1) \\
Turkey         &  2.306(7) &   6.4(1) &  3.2(1) \\
Iran           &  1.862(7) &  5.04(9) &  2.7(1) \\
US             &  2.014(8) &   5.3(1) &  2.9(1) \\
China          &  3.155(8) &   9.4(2) &  3.6(2) \\
\bottomrule
\end{tabular}
